Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-28 doc/invt.tex
replace the thermostat

In practice, however, we may wish to optimize
relative short simulations that have not entered
the asymptotic regime.
%
Then the residual error, $\Err_R$, is not negligible.
%
From Eq. \eqref{eq:error_res}, we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k }
\label{eq:error_res_invt}
\end{equation}
%
where we have assumed that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  E^f_k \,
  \left\{
    1
    +
    \frac{ 1 } { 2 \, \nu_k - 1 }
    \left[
      1 -
      \left(
        \frac{   t_0   }
             { t + t_0 }
      \right)^{2 \, \nu_k - 1}
    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
E^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {  2 \, (t + t_0)      }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t)$.

Two remarks on Eq. \eqref{eq:error_invt} are in order.
%
First, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
since the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$.
%
Second, in the optimal case,
$\nu_k = 1$,
and the final error is twice $E_\mathrm{sat}$
at long times.

@@

For the last part,
we have four roots of
$(z^2 + 1)^2 = - 4 \, \delta^2 \, z^2$,
with two of them lying in the unit circle:
$$
z = \pm i \left( \sqrt{ 1 + \delta^2 } - \delta \right),
$$
such that
$$
z^2 + 1 + 2 \, \delta^2 = 2 \, \delta \, \sqrt{1 + \delta^2}.
$$
Then,
$$
\begin{aligned}
z^2 - \mu_1 (1 + z^2)^2
&= z^2 \, (1 + 4 \, \mu_1 \, \delta^2),
\\
4 \, z^2 - (1 + z^2)^2
&=
4 \, z^2 \, (1 + \delta^2)
\\
%
(2 - \lambda) \, z^2
- 2 \, \mu_1 (1 + z^2)^2
&=
\left(
  2 - \lambda + 8 \, \mu_1 \, \delta^2
\right) \, z^2
\\
%
4 \, z \, (1 + z^2)
+
8 \, \delta^2 \, z
&=
4 \, z \, (z^2 + 1 + \delta^2)
=
4 \, z \, 2 \, \delta \, \sqrt{ 1 + \delta^2 }.
\end{aligned}
$$
Note that the two roots within the unit share the same
residue and the sum for this part is
$$
\operatorname{Res}_3
=
\frac{ (1 + 4 \, \mu_1 \, \delta^2 )^2 }
     { (2 + 8 \, \mu_1 \, \delta^2 - \lambda }
\frac{ \sqrt{ 1 + \delta^2 } }
     {        \delta         }.
$$

>>>>

In practice, however, we may wish to optimize
relative short simulations that have not entered
the asymptotic regime.
%
Then the residual error, $\Err_R$, is not negligible.
%
From Eq. \eqref{eq:error_res}, we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k }
\label{eq:error_res_invt}
\end{equation}
%
where we have assumed that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t)$.

\note{Derivation of Eq. \eqref{eq:error_invt}:
$$
\begin{aligned}
\Err_R
&=
\sum_{ k = 2 }^n
  \Err^f_k \left( \frac{   t_0   }
                       { t + t_0 }
          \right)^{ 2 \nu_k - 1 }
\\
\Err_A
&=
\sum_{ k = 2 }^n
  \frac{ 2 \, \Err^f_k \, \nu_k }
       {    2 \, \nu_k - 1      }
  \left[
    1 -
    \left(
      \tfrac{   t_0   }
            { t + t_0 }
      \right)^{2 \, \nu_k - 1}
  \right]
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left[
    1 -
    \left(
      \tfrac{   t_0   }
            { t + t_0 }
      \right)^{2 \, \nu_k - 1}
  \right]
  +
  \frac{   \Err^f_k     }
       { 2 \, \nu_k - 1 }
  \left[
    1 -
    \left(
      \tfrac{   t_0   }
           { t + t_0 }
      \right)^{2 \, \nu_k - 1}
  \right].
\end{aligned}
$$
Adding the two yields Eq. \eqref{eq:error_invt}.
}

We give two remarks on Eq. \eqref{eq:error_invt}.
%
First, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$.
%
Second, in the optimal case,
$\nu_k = 1$,
and the final error is twice $E_\mathrm{sat}$
at long times.

@@

For the last part,
we have four roots of
$(z^2 + 1)^2 = - 4 \, \delta^2 \, z^2$,
with two of them lying in the unit circle:
$$
z = \pm i \left( \sqrt{ 1 + \delta^2 } - \delta \right),
$$
such that
$$
z^2 + 1 + 2 \, \delta^2 = 2 \, \delta \, \sqrt{1 + \delta^2}.
$$
Then,
$$
\begin{aligned}
z^2 - \mu_1 (1 + z^2)^2
&= z^2 \, (1 + 4 \, \mu_1 \, \delta^2),
\\
4 \, z^2 - (1 + z^2)^2
&=
4 \, z^2 \, (1 + \delta^2)
\\
%
(2 - \lambda) \, z^2
- 2 \, \mu_1 (1 + z^2)^2
&=
\left(
  2 - \lambda + 8 \, \mu_1 \, \delta^2
\right) \, z^2
\\
%
4 \, z \, (1 + z^2)
+
8 \, \delta^2 \, z
&=
4 \, z \, (z^2 + 1 + \delta^2)
=
8 \, z \, \delta \, \sqrt{ 1 + \delta^2 }.
\end{aligned}
$$
Note that the two roots within the unit share the same
residue and the sum for this part is
$$
\operatorname{Res}_3
=
\frac{ (1 + 4 \, \mu_1 \, \delta^2 )^2 }
     { (2 + 8 \, \mu_1 \, \delta^2 - \lambda }
\frac{ \sqrt{ 1 + \delta^2 } }
     {        \delta         }.
$$
